Discovering governing reactions from concentration data

### 2017-11-02 docs/manuscript.tex
wip: manuscript

	Here we demonstrate that, with decreasing time step $\Delta t$ and $\alpha=0$, the estimated rates converge to the true rates if there is no ambiguity in the system. Let there be two species $A$ and $B$ which can unimolecularly convert into each other with macroscopic rates of $k_1=4.0$ and $k_2=0.5$, respectively. The initial amount of particles is set to $70$ of type $A$ and $0$ of type $B$. The to this system corresponding law of mass action solution and a Gillespie realization are  depicted in Figure \ref{ex1:concentration-curves}.

@@

	\subsection{Regression with regularization}
	\label{sec:ex2}
	In this example there are four species $A$, $B$, $C$, and $D$ with reactions $A\xrightarrow{k_1}D$, $D\xrightarrow{k_2}A$, $D\xrightarrow{k_3}B$, $A+B\xrightarrow{k_4}C$, and $C\xrightarrow{k_5}D+B$, where $k_1=k_2=2\cdot 10^{-2}$, $k_3 = 1\cdot 10^{-2}$, and $k_4=k_5=1\cdot 10^{-4}$. These reactions are inserted into the ansatz library $\Theta (\cdot)$. Additionally the reactions $A\leftrightarrow B$, $A\leftrightarrow C$, and $A+C\leftrightarrow D$ are inserted into the library.

@@

	The problem now is to find a suitable $\alpha$ such that a parsimonious solution is yielded. To this end, cross-validation is applied. The trajectories that are used as training and test data set are depicted in Figure \ref{ex2:test-train-concentration-curves}. \textcolor{red}{Perhaps use several test trajectories with different initial conditions to get something like a confidence interval?} As an indicator to the estimation error one can use the scoring function
	\begin{align}
		f_\alpha = \frac{1}{T}\|\mathbf{\dot{X}}_\text{test} - \Theta (\textbf{X}_\text{test}) \Xi_\text{train}\|_F^2,
		\label{ex2:score-function}
	\end{align}
	where the best possible score for some $f_\alpha = 0$ and larger values indicate worse models.
	For a selection of hyperparameters $\alpha\in [0, 100]$, the model parameters $\Xi$ are estimated and the scoring function (\ref{ex2:score-function}) evaluated, yielding a dependency of the score on $\alpha$ as depicted in Figure \ref{ex2:score-graph}.

@@

	One can see that the minimum score is achieved at
	\begin{align}
		\hat{\alpha} = \underset{\alpha\in[0,100]}{\arg\min}f_\alpha\approx 10.
	\end{align}
	When evaluating the minimization problem (\ref{method:minimizationproblem}) with $\hat{\alpha}$ as hyperparameter for the regularizer and applying a cut-off of $10^{-10}$ to the yielded parameters, one obtains the simpler set of reactions $A\xrightarrow{\hat{k}_1}D$, $D\xrightarrow{\hat{k}_3} B$, and $A+B\xrightarrow{\hat{k}_4}C$ with rates $\hat{k}_1 \approx 3.5\cdot 10^{-3}$, $\hat{k}_3\approx 9.1\cdot 10^{-3}$, and $\hat{k}_4\approx 8.7\cdot 10^{-5}$, respectively.

@@

	\begin{figure}
		\begin{center}
			\includegraphics[width=.5\textwidth]{./figures_tex/cv_expected_estimated_concentrations}
		\end{center}
		\caption{Estimated and generating concentration curves as well as the corresponding Gillespie realization for the example in Section \ref{sec:ex2}.}
		\label{ex2:result}
	\end{figure}

>>>>

	Here we demonstrate that, with decreasing time step $\Delta t$ and $\alpha=0$, the estimated rates converge to the true rates if there is no ambiguity in the system. Let there be two species $A$ and $B$ which can unimolecularly convert into each other with macroscopic rates of $k_1=4.0$ and $k_2=0.5$, respectively. The initial amount of particles is set to $70$ of type $A$ and $0$ of type $B$. The law of mass action solution and Gillespie realization corresponding to this system are depicted in Figure \ref{ex1:concentration-curves}.

@@

	\subsection{Regression with regularization}
	\label{sec:ex2}
	In this example there are four species $A$, $B$, $C$, and $D$. The system was generated with reactions $A\xrightarrow{k_1}D$, $D\xrightarrow{k_2}A$, $D\xrightarrow{k_3}B$, $A+B\xrightarrow{k_4}C$, and $C\xrightarrow{k_5}D+B$, where the rates are given in Table \ref{tab:estimated-rates}. These reactions are inserted into the ansatz library $\Theta (\cdot)$. Additionally the reactions $A\leftrightarrow B$, $A\leftrightarrow C$, and $A+C\leftrightarrow D$ are inserted into the library.

@@

	The problem now is to find a suitable $\alpha$ such that a parsimonious solution is yielded. To this end, cross-validation is applied. The trajectories that are used as training and test data set are depicted in Figure \ref{ex2:test-train-concentration-curves}. \textcolor{red}{Perhaps use several test trajectories with different initial conditions to get something like a confidence interval?} As an indicator to the estimation error one can use the scoring function
	\begin{align}
		f_\alpha = \frac{1}{T}\|\mathbf{\dot{X}}_\text{test} - \Theta (\textbf{X}_\text{test}) \Xi_\text{train}\|_F^2,
		\label{ex2:score-function}
	\end{align}
	where the best possible score is $f_\alpha = 0$ and larger values indicate worse models.
	For a selection of hyperparameters $\alpha\in [0, 100]$, the model parameters $\Xi$ are estimated and the scoring function (\ref{ex2:score-function}) evaluated, yielding a dependency of the score on $\alpha$ as depicted in Figure \ref{ex2:score-graph}.

@@

	One can see that the minimum score is achieved at
	\begin{align}
		\hat{\alpha} = \underset{\alpha\in[0,100]}{\arg\min}f_\alpha\approx 10.
	\end{align}
	When evaluating the minimization problem (\ref{method:minimizationproblem}) with $\hat{\alpha}$ as hyperparameter for the regularizer and applying a cut-off of $10^{-10}$ to the yielded parameters, one obtains the simpler set of reactions $A\xrightarrow{\hat{k}_1}D$, $D\xrightarrow{\hat{k}_3} B$, and $A+B\xrightarrow{\hat{k}_4}C$ with rates as in Table \ref{tab:estimated-rates}.

	\begin{table}
		\centering
		\begin{tabular}{ccc}
			Reaction & original rate & estimated rate \\ \hline\noalign{\vskip .1cm}
			$A \to D$ & $k_1 = 2\cdot 10^{-2}$ & $\hat{k}_1 \approx 3.5\cdot 10^{-3}$\\
			$D \to A$ & $k_2 = 2\cdot 10^{-2}$ & $\hat{k}_2 = 0$ \\
			$D\to B$ & $k_3 = 1\cdot 10^{-2}$ & $\hat{k}_3\approx 9.1 \cdot 10^{-3}$ \\
			$A+B\to C$ & $k_4 = 1\cdot 10^{-4}$ & $\hat{k}_4\approx 8.7 \cdot 10^{-5}$ \\
			$C \to D+B$ & $k_5 = 1\cdot 10^{-4}$ & $\hat{k}_5 = 0$ 
		\end{tabular}
		\caption{Original and estimated reaction rates for the example in Section \ref{sec:ex2}.}
		\label{tab:estimated-rates}
	\end{table}

@@

	\begin{figure}
		\begin{center}
			\includegraphics[width=.5\textwidth]{./figures_tex/cv_expected_estimated_concentrations}
		\end{center}
		\caption{Integrated law of mass action equations (\ref{method:the-system}) for the sparse estimated parameters and the corresponding Gillespie realization for the example in Section \ref{sec:ex2}.}
		\label{ex2:result}
	\end{figure}
